The Strong Gravitational Lens Finding Challenge

### 2017-08-16 paper.tex
Update on Overleaf.

Strong gravitational lenses are rare cases where a distant galaxy or quasar is aligned so closely with a foreground galaxy or cluster of galaxies  that the gravitational field of the foreground object creates multiple, highly distorted images of the background object.  These take the form of multiple images of a quasar and Einstein rings surrounding a galaxy and of highly distorted arcs seen through galaxy groups and clusters.  The first strong lens was discovered in 1979 \citep{1979Natur.279..381W} and since then many valuable scientific uses have been found for them.  They have been used to study how dark matter is distributed in galaxies and clusters \citep[e.g.][]{1991ApJ...373..354K,2001ApJ...554.1216C,2002ApJ...568L...5K,2003ApJ...587..143R,2003ApJ...583..606K,2005MNRAS.360.1333W,2005ApJ...623...31D,2009MNRAS.392..945V,2016MNRAS.463.3115T}.  They have been used to measure the Hubble constant and other cosmological parameters \citep[e.g.][]{1964MNRAS.128..307R,1992ARA&A..30..311B,2000ApJ...544...98W,2013ApJ...766...70S,2016A&ARv..24...11T}.
Their magnification has been used as a natural telescope to observe otherwise undetectable objects \citep[e.g][]{2007ApJ...671.1196M,2017MNRAS.464.4823B,2016ApJ...833..264S}.   They have put limits on the self interaction of dark matter and on the theory of gravitation.  Through microlensing they have been used to study the structure of quasars.   To access this wealth of information the first step is to find the lenses.  \red{Could use some citations in this paragraph.}

@@

Up to this point, the most widely used method for finding lenses in imaging surveys has been by visual inspection of a samples candidates that have been selected on the bases of luminosity and/or color.  This has been done in the radio \citep{2003MNRAS.341...13B} and in the visible with space and ground based data \citep{2008MNRAS.389.1311J,2008ApJS..176...19F,2014MNRAS.439.3392P} \red{(Maybe there are some others to cite here?)}.  A related method pioneered by the SPACE WARPS project \citep{2016MNRAS.455.1171M,2016MNRAS.455.1191M} has been to crowd source the visual inspection through an online platform.  These efforts have proved very fruitful, but they deal with orders of magnitude fewer candidates and lenses than will be necessary in the future.  Dealing with such large quantities of data will not be practical for any visual inspection approach.  In addition, the efficiency and detection bias of human inspection methods are difficult to rigorously quantify.

Galaxy scale lenses have also been searched for spectroscopicly by looking for high redshift stellar lines in the spectra of lower redshift large galaxies.  Notably this was done in the Sloan Lens ACS (SLACS) survey to producing a relatively well defined and pure sample of Einstein ring lenses \citep{2006ApJ...638..703B}.  However, spectroscopy is telescope-time consuming and for the foreseeable future we are not likely to have spectroscopic surveys that cover anywhere near the number of objects as the planned imaging surveys.

@@

More recently machine learning techniques that have become widely used in the fields computer image processing and artificial intelligence 
have been applied to this problem and others in astronomy, in particular, artificial neural networks (ANNs) and support vector machines (SVM). These algorithms do not pick out particular predefined features, but learn from a training set which features are most important for distinguishing categories of objects.  These have been used widely for such tasks as handwriting and facial recognition.  In astronomy they are beginning to be used for catagorizing galaxies morphologies \citep{2015MNRAS.450.1441D}, photometric redshifts \citep{2017MNRAS.465.1959C,2016PASP..128j4502S,2017NewA...51..169S}, supernova classification \citep{2016ApJS..225...31L} and the lens finding problem \citep{2017MNRAS.471..167J,2017arXiv170207675P,2017MNRAS.465.4325O}.

Given the future of this field with large amounts of data coming soon and many new ideas emerging, it is timely to stage a series of challenges to stimulate new work, determine what can realistically be done in lens finding and get a better idea of the strengths and weaknesses to different methods.  The long term goal is to get a set of algorithms that can handle Euclid, LSST or SKA data sets and produce high purity and high completeness lens samples with well defined efficiency or selection.  We anticipate further challenges in the future were the realism of the data simulations will become progressively larger.  Here we have chosen to concentrate on galaxy/small group scale lenses where the background source is a galaxy because we feel that this is where the most progress can be made and the scientific return is the highest, but QSO lenses and cluster/group lenses challenges may follow.

@@

These foreground galaxies are rotated in three dimensions with the halos of the lens each time a random lens is produced so that they remain in the same same positions relative to the mass.  All the random parameters are also reassigned with every realization of the lens.

@@

The KiDS survey provided a representative PSF map in each band that was applied to all mock images.   The pixel size in this case was 0.2~arcsec. 
Weight maps for the KiDS images were also provided.  Some of these had masked regions from removed stars, cosmic rays, and bad pixels.  For the mock images the noise was simulated by adding normally distributed numbers with the variance given by the weight maps.  The weight maps where also randomly rotated and flipped.  This resulted in many of the images having large masked regions in them.

@@

The Python scripts used for manual examination of multiple images are available
onGitHub\footnote{https://github.com/nealjackson/bigeye} and are described in more detail
in Hartley et al. (2016, in preparation). For one-colour data such as
the space-based training set, the images are individually colour-scaled using
square-root scaling. The bright limit of the colour-scale is determined
from the pixel values in a rectangle comprising the inner ninth of the
image area, with the limit being chosen as the $n$th centile of the pixel
values in this area. Values between $n=95$ and $n=98$ give optimum results,
judging by experiments on the training set. The number of images in each
grid was also optimised using the training set, with 16$\times$8 or 
8$\times$4 giving good results on one-colour data. For three-colour data,
such as the ground-based challenge data, the individual bands for each 
object are colour-scaled and then combined into an RGB image. In this case
8$\times$4 grids were used for examination, due to the generally lower 
resolution of the images. The script also allows the user to adjust the
colour-scale in real time when examining and marking images, and records
the image name corresponding to the image within which the cursor resides
at the time any key is pressed, together with the key.

>>>>

Strong gravitational lenses are rare cases where a distant galaxy or quasar is aligned so closely with a foreground galaxy or cluster of galaxies  that the gravitational field of the foreground object creates multiple, highly distorted images of the background object.  These take the form of multiple images of a quasar and Einstein rings surrounding a galaxy and of highly distorted arcs seen through galaxy groups and clusters.  The first strong lens was discovered in 1979 \citep{1979Natur.279..381W} and since then many valuable scientific uses have been found for them.  They have been used to study how dark matter is distributed in galaxies and clusters \citep[e.g.][]{1991ApJ...373..354K,2001ApJ...554.1216C,2002ApJ...568L...5K,2003ApJ...587..143R,2003ApJ...583..606K,2005MNRAS.360.1333W,2005ApJ...623...31D,2009MNRAS.392..945V,2016MNRAS.463.3115T}.  They have been used to measure the Hubble constant and other cosmological parameters \citep[e.g.][]{1964MNRAS.128..307R,1992ARA&A..30..311B,2000ApJ...544...98W,2013ApJ...766...70S,2016A&ARv..24...11T}.
Their magnification has been used as a natural telescope to observe otherwise undetectable objects \citep[e.g][]{2007ApJ...671.1196M,2017MNRAS.464.4823B,2016ApJ...833..264S}.   They have put limits on the self interaction of dark matter and on the theory of gravitation.  Through microlensing they have been used to study the structure of quasars \citep{2008ApJ...689..755M,2008ApJ...673...34P,2011ApJ...729...34B}.   To access this wealth of information the first step is to find the lenses.  \red{Could use some citations in this paragraph.}

@@

Up to this point, the most widely used method for finding lenses in imaging surveys has been by visual inspection of a samples candidates that have been selected on the bases of luminosity and/or color.  This has been done in the radio \citep{2003MNRAS.341...13B} and in the visible with space and ground based data \citep{2008MNRAS.389.1311J,2008ApJS..176...19F,2014MNRAS.439.3392P} \red{(Maybe there are some others to cite here?)}.  A related method pioneered by the SPACE WARPS project \citep{2016MNRAS.455.1171M,2016MNRAS.455.1191M,2015MNRAS.452..502G} has been to crowd source the visual inspection through an online platform.  These efforts have proved very fruitful, but they deal with orders of magnitude fewer candidates and lenses than will be necessary in the future.  Dealing with such large quantities of data will not be practical for any visual inspection approach.  In addition, the efficiency and detection bias of human inspection methods are difficult to rigorously quantify.

Galaxy scale lenses have also been searched for spectroscopically by looking for high redshift stellar lines in the spectra of lower redshift large galaxies.  Notably this was done in the Sloan Lens ACS (SLACS) survey, producing a relatively well defined and pure sample of Einstein ring lenses \citep{2006ApJ...638..703B}.  However, spectroscopy is telescope-time consuming and for the foreseeable future we are not likely to have spectroscopic surveys that cover anywhere near the number of objects as the planned imaging surveys.

@@

More recently machine learning techniques that have become widely used in the fields computer image processing and artificial intelligence 
have been applied to this problem and others in astronomy, in particular, artificial neural networks (ANNs) and support vector machines (SVM). These algorithms do not pick out particular predefined features, but learn from a training set which features are most important for distinguishing categories of objects.  These have been used widely for such tasks as handwriting and facial recognition.  In astronomy they are beginning to be used for categorizing galaxy morphologies \citep{2015MNRAS.450.1441D}, photometric redshifts \citep{2017MNRAS.465.1959C,2016PASP..128j4502S,2017NewA...51..169S}, supernova classification \citep{2016ApJS..225...31L} and the lens finding problem \citep{2017MNRAS.471..167J,2017arXiv170207675P,2017MNRAS.465.4325O}.

Given the future of this field with large amounts of data coming soon and many new ideas emerging, it is timely to stage a series of challenges to stimulate new work, determine what can realistically be done in lens finding and get a better idea of the strengths and weaknesses to different methods.  The long term goal is to get a set of algorithms that can handle Euclid, LSST or SKA data sets and produce high purity and high completeness lens samples with well defined efficiency or selection.  We anticipate further challenges in the future in which the realism of the data simulations will become progressively better.  Here we have chosen to concentrate on galaxy/small group scale lenses where the background source is a galaxy because we feel that this is where the most progress can be made and the scientific return is the highest, but QSO lenses and cluster/group lenses challenges may follow.

@@

These foreground galaxies are rotated in three dimensions with the halos of the lens each time a random lens is produced so that they remain in the same positions relative to the mass.  All the random parameters are also reassigned with every realization of the lens.

@@

The KiDS survey provided a representative PSF map in each band that was applied to all mock images.   The pixel size in this case was 0.2~arcsec. 
Weight maps for the KiDS images were also provided.  Some of these had masked regions from removed stars, cosmic rays, and bad pixels.  For the mock images the noise was simulated by adding normally distributed numbers with the variance given by the weight maps.  The weight maps were also randomly rotated and flipped.  This resulted in many of the images having large masked regions in them.

@@

The Python scripts used for manual examination of multiple images are available
onGitHub\footnote{https://github.com/nealjackson/bigeye} and are described in more detail
in \citet{hartley2017support}. For one-colour data such as
the space-based training set, the images are individually colour-scaled using
square-root scaling. The bright limit of the colour-scale is determined
from the pixel values in a rectangle comprising the inner ninth of the
image area, with the limit being chosen as the $n$th centile of the pixel
values in this area. Values between $n=95$ and $n=98$ give optimum results,
judging by experiments on the training set. The number of images in each
grid was also optimised using the training set, with 16$\times$8 or 
8$\times$4 giving good results on one-colour data. For three-colour data,
such as the ground-based challenge data, the individual bands for each 
object are colour-scaled and then combined into an RGB image. In this case
8$\times$4 grids were used for examination, due to the generally lower 
resolution of the images. The script also allows the user to adjust the
colour-scale in real time when examining and marking images, and records
the image name corresponding to the image within which the cursor resides
at the time any key is pressed, together with the key.
